<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Units of Measurement for R Vectors: an Introduction}
-->

# Units of Measurement for R Vectors: an Introduction

R has little support for physical measurement units.  The exception
is formed by time differences: time differences objects of class
`difftime` have a `units` attribute that can be modified:

```{r}
t1 = Sys.time() 
t2 = t1 + 3600 
d = t2 - t1
class(d)
units(d)
d
units(d) = "secs"
d
```
We see here that the `units` method is used to retrieve and modify the
unit of time differences. 

This idea can be generalized to other physical units. The
`units` package, presented here, does this, and builds upon the
[udunits2](https://cran.r-project.org/package=udunits2)
R package, which in turn is build upon the
[udunits2](https://www.unidata.ucar.edu/software/udunits/) C library.
The `udunits2` library provides the following operations:

* validating whether a string, such as `m/s` is a valid physical unit
* verifying whether two strings such as `m/s` and `km/h` are convertible
* converting values between two convertible units 
* providing names and symbols for specific units
* handle different character encodings (utf8, ascii, iso-8859-1 and latin1)

Th `units` R package uses R package udunits2 to extend
R with functionality for manipulating numeric vectors that have
physical measurement units associated with them, in a similar way as
`difftime` objects behave.

## Setting units, unit conversion

Units must be defined before they are used. Units are defined from a string but by explicitly defining units before they are used we avoid having new units in use due to misspellings. A unit is defined using the `make_units` function

```{r}
library(units)
m <- make_unit("m")
s <- make_unit("s")
```
The objects created, 
```{r}
m
```
represent a single value of the named quantity (``m'').  To obtain
different quantities of named units, we use them in expressions
such as 
```{r}
(a <- 1:10 * m/s)
```
the result, e.g. 
```{r}
a[10]
```
literally means "10 times 1 m divided by 1 s". In writing, the "1"
values are omitted, and the multiplication is implicit.

The `units` package comes with a list of over 3000 predefined units, which
can be used without calling `make_unit`. 
One way of making this list available would be to attach it, but
this leads to a number of name conflicts: `T` (for tesla) wound
mask the value that otherwise refers to `TRUE`, .

A cleaner approach is to use `with`, as in
```{r}
rm(m) ; rm(s)
with(ud_units, 1:10 * m/s)
```

When conversion is meaningful, such as hours to seconds or meters to kilometers, conversion can be done explicitly by setting the units of a vector
```{r}
b = a
units(b) <- with(ud_units, km/h)
b
```

## Basic manipulations

### Arithmetic operations

Arithmetic operations verify units, and create new ones
```{r}
a + a
a * a
a ^ 2
a ** -2
```
and convert to the units of the first argument if necessary:
```{r}
a + b # m/s + km/h -> m/s
```

Currently, powers are only supported for integer powers, so using `a ** 2.5` would result in an error.

There are some basic simplification of units:
```{r}
t <- with(ud_units, s)
a * t
```
which also work when units need to be converted before they can be simplified:
```{r}
t <- with(ud_units, min)
a * t
```
Simplification to unit-less values gives the "1" as unit:
```{r}
m <- with(ud_units, m)
a * t / m
```

Allowed operations that require convertible units are `+`, `-`, `==`,
`!=`, `<`, `>`, `<=`, `>=`.  Operations that lead to new units are
`*`, `/`, and the power operations `**` and `^`.


### Mathematical functions

Mathematical operations allowed are: `abs`, `sign`, `floor`,
`ceiling`, `trunc`, `round`, `signif`, `log`, `cumsum`, `cummax`, `cummin`.
```{r}
signif(a ** 2 / 3, 3)
cumsum(a)
log(a) # base defaults to exp(1)
log(a, base = 10)
log(a, base = 2)
```

### Summary functions
Summary functions `sum`, `min`, `max`, and `range` are allowed:
```{r}
sum(a)
min(a)
max(a)
range(a)
with(ud_units, min(m/s, km/h)) # converts to first unit:
```

### Printing
Following `difftime`, printing behaves differently for length-one vectors:
```{r}
a
a[1]
```

### Subsetting
The usual subsetting rules work:
```{r}
a[2:5]
a[-(1:9)]
```

### Concatenation
```{r}
c(a,a)
```
concatenation converts to the units of the first argument, if necessary:
```{r}
c(a,b) # m/s, km/h -> m/s
c(b,a) # km/h, m/s -> km/h
```

## Conversion to/from `difftime`
From `difftime` to `units`:
```{r}
t1 = Sys.time() 
t2 = t1 + 3600 
d = t2 - t1
du <- as.units(d)
```
vice versa:
```{r}
dt = as.dt(du)
class(dt)
dt
```

## units in `matrix` objects
```{r}
with(ud_units, matrix(1:4,2,2) * m/s)
with(ud_units, matrix(1:4,2,2) * m/s * 4 * m/s)
```
but
```{r}
with(ud_units, (matrix(1:4,2,2) * m/s) %*% (4:3 * m/s))
```
strips units.

## units objects in `data.frame`s
units in `data.frame` objects are printed, but do not appear in `summary`:.
```{r}
set.seed(131)
d <- with(ud_units,
         data.frame(x = runif(4), 
                    y = runif(4) * s, 
                    z = 1:4 * m/s))
d
summary(d)
d$yz = with(d, y * z)
d
d[1, "yz"]
```
## formatting

Units are often written in the form `m2 s-1`, for square meter per second. This
can be defined as unit, but is not interpreted by R:
```{r}
(x = 1:10 * make_unit("m2 s-1"))
```
udunits understands such string, and can convert them
```{r}
y = 1:10 * with(ud_units, m^2/s)
x + y
```
but R cannot simplify them:
```{r}
x/y
```
Instead, we can tell R to parse such a string, which then allows simplification:
```{r}
(z = 1:10 * parse_unit("m2 s-1"))
z + y
z / y
```
Printing units in this form is done by
```{r}
as_cf(z)
```

## plotting

Base scatter plots and histograms support automatic unit placement
in axis labels. In the following example we first convert to
SI units. (Unit `in` needs a bit special treatment, because `in` is a
reserved word in R.)

```{r fig=TRUE}
mar = par("mar") + c(0, .3, 0, 0)
displacement = mtcars$disp * ud_units[["in"]]^3
units(displacement) = with(ud_units, cm^3)
weight = mtcars$wt * 1000 * with(ud_units, lb)
units(weight) = with(ud_units, kg)
par(mar = mar)
plot(weight, displacement)
```

We can change grouping symbols from `[ ]` into `( )`:
```{r}
units_options(group = c("(", ")") )  # parenthesis instead of square brackets
par(mar = mar)
plot(weight, displacement)
```

We can also remove grouping symbols, increase space between variable name and unit by:
```{r}
units_options(sep = c("~~~", "~"), group = c("", ""))  # no brackets; extra space
par(mar = mar)
plot(weight, displacement)
```

More complex units can be plotted either with negative powers, or as divisions,
by modifying one of `units`'s global options using `units_options`:

```{r}
gallon = make_unit("gallon")
consumption = mtcars$mpg * with(ud_units, mi/gallon)
units(consumption) = with(ud_units, km/l)
par(mar = mar)
plot(displacement, consumption) # division in consumption
units_options(negative_power = TRUE) # division becomes ^-1
plot(displacement, consumption) # division in consumption
```

As usual, units modify automatically in expressions:

```{r}
units_options(negative_power = TRUE) # division becomes ^-1
par(mar = mar)
plot(displacement, consumption)
plot(1/displacement, 1/consumption)
```
